---
title: "Week 2: Data Wrangling & Exploratory Ecological Analysis"
subtitle: "ENV3QRM — Quantitative Research Methods"
author: "La Trobe University"
date: "Semester 1, 2026"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    number_sections: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

This workshop builds on the R basics from Week 1. You will work with a larger
ecological monitoring dataset (`eco_monitor`) to practise:

- Selecting, filtering and counting observations.
- Standardising abundance by sampling effort.
- Computing grouped summaries across habitats and months.
- Building scatter plots, boxplots and line graphs with **ggplot2**.
- Preparing a tidy dataset ready for statistical modelling.

**Estimated time:** approximately 1 hour.

## How to use this worksheet

1. Open R or RStudio with `hello-eco-stats-r` as your working directory.
2. Source the launcher script to start **swirl**:

```{r}
source("01_introduction_to_r_programming.R")
```

3. In swirl, choose **Intro_Ecological_Statistics_in_R** →
   **02_data_wrangling_and_exploratory_ecological_analysis**.
4. Work through each question below in the swirl console. Use this document as
   your reference — copy each question into the R source pane, write your
   answer beneath it, and run it.

---

## The dataset: `eco_monitor`

`eco_monitor` is pre-loaded when the lesson starts. It contains monthly
ecological monitoring data from three sites across four months (Jan–Apr).

| Column             | Description                             |
|--------------------|-----------------------------------------|
| `month`            | Survey month                            |
| `site`             | Site code (Wetland_A, Forest_C, Grass_D)|
| `habitat`          | Habitat type                            |
| `species_richness` | Number of species recorded              |
| `abundance`        | Total individuals counted               |
| `temp_c`           | Air temperature (°C)                    |
| `rainfall_mm`      | Rainfall (mm)                           |
| `effort_hours`     | Hours spent surveying                   |

---

## Section 1: Loading Packages & Inspecting Data

### Background

Before any analysis, load the packages you need and inspect your data. Always
check the structure and first rows so you understand what you are working with
before writing any analysis code.

### Questions

Load the tidyverse package (swirl will prompt you).

```{r}
library(tidyverse)
```

**Q1.** Preview the `eco_monitor` dataset using `head()`.

```{r}
head(eco_monitor)
```

**Q2.** Check the structure of `eco_monitor` using `str()`.

```{r}
str(eco_monitor)
```

**Q3.** Which variable in `eco_monitor` describes sampling effort?

- effort_hours
- month
- species_richness
- rainfall_mm

---

## Section 2: Selecting & Filtering

### Background

In larger datasets you often need only a subset of columns or rows. Use
`select()` to choose columns and `filter()` to keep rows matching a condition.
These are your first two moves when wrangling ecological data.

### Questions

**Q4.** Select only `month`, `site`, `habitat`, `abundance`, and `effort_hours`
columns.

```{r}
eco_monitor %>% select(month, site, habitat, abundance, effort_hours)
```

**Q5.** Create `wet_season` as rows with `rainfall_mm >= 100`.

```{r}
wet_season <- eco_monitor %>% filter(rainfall_mm >= 100)
```

**Q6.** Print `wet_season`.

```{r}
wet_season
```

**Q7.** Create `grass_data` containing only `habitat == "grassland"`.

```{r}
grass_data <- eco_monitor %>% filter(habitat == "grassland")
```

**Q8.** Count how many observations occur in each month.

```{r}
eco_monitor %>% count(month)
```

---

## Section 3: Effort Standardisation

### Background

Ecological surveys often vary in the time spent sampling. Comparing raw counts
between a 5-hour survey and a 7-hour survey is misleading. **Effort
standardisation** divides abundance by effort (e.g. hours) so that comparisons
are fair. This is standard practice in wildlife monitoring.

### Questions

**Q9.** Create `effort_standardized` by adding
`abundance_per_hour = abundance / effort_hours`.

```{r}
effort_standardized <- eco_monitor %>%
  mutate(abundance_per_hour = abundance / effort_hours)
```

**Q10.** Display `site`, `month`, and `abundance_per_hour` from
`effort_standardized`.

```{r}
effort_standardized %>% select(site, month, abundance_per_hour)
```

> *Why this matters:* standardising by effort makes comparisons fairer across
> surveys of different durations — a fundamental step before ecological
> inference.

---

## Section 4: Grouped Summaries

### Background

Grouped summaries are the workhorse of ecological data exploration. By applying
`group_by()` + `summarise()` you can answer questions like *"What is the mean
abundance in each habitat?"* or *"How does species richness change from month to
month at each site?"*. You can group by multiple variables at once.

### Questions

**Q11.** Summarise mean abundance by habitat (name the result
`habitat_mean_abundance`).

```{r}
habitat_mean_abundance <- eco_monitor %>%
  group_by(habitat) %>%
  summarise(mean_abundance = mean(abundance))
```

**Q12.** Print `habitat_mean_abundance`.

```{r}
habitat_mean_abundance
```

**Q13.** Calculate mean `species_richness` by `site` and `month`.

```{r}
eco_monitor %>%
  group_by(site, month) %>%
  summarise(mean_richness = mean(species_richness))
```

**Q14.** Create `habitat_month_summary` with `mean_abundance` and
`mean_richness` by `habitat` and `month`.

```{r}
habitat_month_summary <- eco_monitor %>%
  group_by(habitat, month) %>%
  summarise(
    mean_abundance = mean(abundance),
    mean_richness  = mean(species_richness)
  )
```

**Q15.** Arrange `habitat_month_summary` by descending `mean_abundance`.

```{r}
habitat_month_summary %>% arrange(desc(mean_abundance))
```

**Q16.** Why is grouped summarisation useful in ecology?

- It compares patterns across habitats or time periods
- It guarantees causal conclusions
- It removes all sampling bias
- It replaces field observations

---

## Section 5: Exploratory Visualisation

### Background

Visualisation turns numbers into insight. Three common ecological plot types:

| Plot type    | Use case                                     |
|--------------|----------------------------------------------|
| Scatter plot | Relationship between two continuous variables|
| Boxplot      | Distribution of a variable across groups     |
| Line graph   | Trends over time                             |

ggplot2 layers these elements: **data → aesthetics → geom → (optional)
facets, labels, themes**.

### Questions

**Q17.** Create a scatter plot of `rainfall_mm` vs `abundance` with points
coloured by `habitat`.

```{r}
ggplot(eco_monitor, aes(x = rainfall_mm, y = abundance, color = habitat)) +
  geom_point()
```

**Q18.** Add a linear trend line to that scatter plot (without confidence band).

```{r}
ggplot(eco_monitor, aes(x = rainfall_mm, y = abundance, color = habitat)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

**Q19.** Create a boxplot of `species_richness` by `habitat`.

```{r}
ggplot(eco_monitor, aes(x = habitat, y = species_richness)) +
  geom_boxplot()
```

**Q20.** Create `monthly_trend` with total abundance per month.

```{r}
monthly_trend <- eco_monitor %>%
  group_by(month) %>%
  summarise(total_abundance = sum(abundance))
```

**Q21.** Plot monthly total abundance as a line graph with points.

```{r}
ggplot(monthly_trend, aes(x = month, y = total_abundance, group = 1)) +
  geom_line() +
  geom_point()
```

---

## Section 6: Preparing Data for Modelling

### Background

Before fitting statistical models you typically create a tidy, analysis-ready
dataset. This means selecting only the variables relevant to your research
question, checking for missing values, and exporting the result so your
modelling script can read it in.

### Questions

**Q22.** Create `model_ready` by selecting `month`, `site`, `habitat`,
`species_richness`, `abundance`, `rainfall_mm`, `temp_c`, and `effort_hours`.

```{r}
model_ready <- eco_monitor %>%
  select(month, site, habitat, species_richness,
         abundance, rainfall_mm, temp_c, effort_hours)
```

**Q23.** Write `model_ready` to `"model_ready_ecology.csv"` without row names.

```{r}
write.csv(model_ready, "model_ready_ecology.csv", row.names = FALSE)
```

**Q24.** Read `"model_ready_ecology.csv"` back into `model_ready_check`.

```{r}
model_ready_check <- read.csv("model_ready_ecology.csv")
```

**Q25.** Display the first rows of `model_ready_check`.

```{r}
head(model_ready_check)
```

---

## Section 7: Reflection & Reproducibility

### Questions

**Q26.** After exploratory analysis, what is a good next step in an ecological
statistics workflow?

- Fit and evaluate statistical models based on your questions
- Delete raw data files immediately
- Use only one graph as final evidence
- Stop analysis without documenting methods

**Q27.** Run `sessionInfo()` to record software details for reproducibility.

```{r}
sessionInfo()
```

---

## Summary

In this workshop you learned to:

1. Inspect and explore monitoring data.
2. Select columns and filter rows for targeted analysis.
3. Standardise abundance by sampling effort.
4. Compute grouped summaries across habitats and time periods.
5. Build scatter plots, boxplots and line graphs with ggplot2.
6. Prepare and export a modelling-ready dataset.

**Next steps:** inferential methods in ecology — linear models, generalised
linear models and model diagnostics.
