---
author: Quantitative Research Methods
date: 2026
output:
  html_document:
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
title: Introduction to R
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(klippy)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c("top", "right"), tooltip_message = "Copy code", tooltip_success = "Copied!")
```

## Workshop overview

Welcome to our first R workshop! 

In this workshop you will learn the **fundamentals of R**, a
powerful, free, open-source programming language widely used in life
sciences and data analysis.

By the end of this workshop you will be able to:

-   Use R as a **calculator** and store results in variables
-   Understand **basic data types**, **vectors**, and **data frames**
-   Install and use **R packages** and **functions**
-   Set up a **project structure** and import data
-   Wrangle data using the **tidyverse**
-   Create basic plots using **ggplot2**

------------------------------------------------------------------------

# 1. R as a calculator

One of the simplest ways to start using R is as a **calculator**.

You type an expression, run it, and R returns the result.

## 1.1 Basic arithmetic operations

```{r}
# Addition
3 + 5

# Subtraction
10 - 4

# Multiplication
6 * 7

# Division
20 / 4

# Exponentiation (2 to the power of 3)
2^3
```

### Order of operations

R follows standard mathematical rules (**BODMAS / PEMDAS**).

```{r}
# Without parentheses
2 + 3 * 4

# With parentheses
(2 + 3) * 4
```

Parentheses allow you to **control the order of operations**.

------------------------------------------------------------------------

## 1.2 Assigning values to variables

Instead of calculating a value and losing it, you can **store a value in a variable**.

A variable is simply a **named object that stores a value in the environment**.

We assign values using the **assignment operator** `<-`.

```{r}
x <- 10
y <- 3
```

Now we can use these variables in calculations.

```{r}
x + y
x * y
```

### Updating variables

Variables can be overwritten by assigning a new value.

```{r}
x <- 10
x

x <- 99
x
```

### Naming variables

Good variable names make your code easier to read.

**Rules for variable names**

- Use letters, numbers, `_`, or `.`
- Names must start with a **letter**
- Names are **case-sensitive**. For example, `MyVar` and `myvar` are treated as **different variables**.

### Naming styles

  Style        Example          Description
  ------------ ---------------- ----------------
  snake_case   `plant_height`   Lowercase words separated by underscores
  camelCase    `plantHeight`    Each word capitalised except the first

In R, **snake_case is generally preferred** because it is easier to
read.

------------------------------------------------------------------------

## 1.3 Basic data types in R

Every value in R has a **data type**.

  Type          Description   Example
  ------------- ------------- -----------------
  `numeric`     Numbers       `3.14`, `42`
  `character`   Text          `"hello"`
  `logical`     TRUE/FALSE    `TRUE`, `FALSE`

Check the type of a value using `class()`.

```{r}
class(3.14)
class("hello")
class(TRUE)
```

------------------------------------------------------------------------

## 1.4 Vectors

A **vector** is a collection of values of the same type.

```{r}
heights_cm <- c(105, 150, 112, 80, 132)
heights_cm

genus <- c("Banksia", "Eucalyptus", "Acacia")
genus
```

<br>
R applies mathematical operations **to every element**.

```{r}
heights_cm / 100
```

------------------------------------------------------------------------

## 1.5 Data frames

A **data frame** stores tabular data.

```{r}
plant_data <- data.frame(
  genus = c("Banksia", "Eucalyptus", "Acacia"),
  height_cm = c(150, 300, 200)
)

print(plant_data)
```

<br>
Access a column using `$` or double square brackets `[[ ]]`:

```{r}
plant_data$height_cm
plant_data[["height_cm"]]
```

------------------------------------------------------------------------

# 2. R packages and functions

## 2.1 Functions

A **function** performs a specific task.

Example:

```{r}
mean(c(10, 20, 30))
length(c(10, 20, 30))
```

-   `mean()` calculates the average
-   `length()` counts the number of elements

------------------------------------------------------------------------

## 2.2 Packages

A **package** is a collection of functions that extends what R can do.

------------------------------------------------------------------------

## 2.3 Installing and loading packages

You only need to **install a package once**, but you must **load the package from your package library each session**.

```{r eval=FALSE}
# Install packages
install.packages("tidyverse")
```

```{r, message=FALSE, warning=FALSE}
# Load packages
library(tidyverse)
```

The **tidyverse** is a collection of core R packages designed for data science, including:

-   **dplyr** -- data manipulation
-   **ggplot2** -- visualisation
-   **readr** -- importing data

------------------------------------------------------------------------

## 2.4 Using functions from packages

```{r}
glimpse(plant_data)
```

`glimpse()` gives a **compact overview of a data frame**.

### Function conflicts

If two packages contain the same function name, you can specify the
package:

```{r}
dplyr::glimpse(plant_data)
```
This can be helpful when **different packages contain functions with the same name**.

------------------------------------------------------------------------

# 3. Project structure

Good project organisation improves **reproducibility**.

    my_project/
    ├── my_project.Rproj
    ├── data/
    ├── scripts/
    └── output/

Check your working directory:

```{r}
getwd()
```

<br>
List files:

```{r}
list.files("./data")
```

<br>
Read data:
```{r, message=FALSE, warning=FALSE}
banksia_data <- read_csv(file = "./data/banksia_data.csv")
```
Here, the `read_csv()` looks for the file **banksia_data.csv** in the **data** subdirectory.


------------------------------------------------------------------------

# 4. Data wrangling with the tidyverse

Common **dplyr** functions:

  Function      Purpose
  ------------- ----------------
  `select()`      choose columns
  `filter()`      choose rows
  `mutate()`      create columns
  `summarise()`   summaries
  `arrange()`     sort rows

### Pipes

The pipe `%>%` means **"and then"**.

```{r}
banksia_data %>%
  select(pair, treatment)
```
Here, R reads `banksia_data` ***and then*** selects the columns ***pair*** and ***treatment***

------------------------------------------------------------------------

# 5. Data visualisation with ggplot2

Every ggplot requires:

1.  Data
2.  Aesthetics (`aes()`)
3.  Geometry (`geom_*()`)

### Example boxplot

```{r}
# Create a box plot
banksia_boxplot <- ggplot(
  data = banksia_data, 
  aes(x = treatment, y = height_change_cm)
  ) +
  geom_boxplot()

print(banksia_boxplot)
```

### Example scatter plot

```{r}
# Create a scatter plot
banksia_scatter <- ggplot(
  data = banksia_data,
  aes(x = height_2013_cm, y = height_2016_cm, colour = treatment)
  ) +
  geom_point()

print(banksia_scatter)
```

### Saving plots

```{r eval=FALSE}
# Save the boxplot
ggsave(
  filename = "output/banksia_boxplot.png",
  plot = banksia_boxplot,
  width = 12,
  height = 8,
  units = "cm"
)
```

# 6. Check your understanding

> **Q1.** Run `x <- 5` and then `y <- x + 3` in R. What value does `y` hold?
>
> **Q2.** Check the data type of `x` and `y. Are they numeric or character?
>
> **Q3.** How would you call the `filter()` function from the **dplyr** package without loading the package first?
>
> **Q4.** What does the pipe operator `%>%` do?
>
> **Q5.** Pipe the `banksia_data` into `mutate()` function to recompute `hieght_change_cm` as the difference between `hieght_2016_cm` and `hieght_2013_cm` (Hint: `hieght_change_cm = hieght_2016_cm - hieght_2013_cm`)
>
> **Q6.** Which `geom_*()` function do you think might be used for a ***bar*** chart
>


