---
title: "Week 1: Introduction to R Programming"
subtitle: "ENV3QRM — Quantitative Research Methods"
author: "La Trobe University"
date: "Semester 1, 2026"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    number_sections: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

This workshop introduces the R programming language in the context of ecological
data analysis. By the end of the session you will be able to:
 
- Create and inspect R objects (scalars and vectors).
- Read ecological survey data from a CSV file.
- Perform basic calculations and subset data using base R.
- Use **tidyverse** pipelines to filter, mutate, summarise and arrange data.
- Produce exploratory plots with **ggplot2**.
- Export results to CSV and record your session for reproducibility.

**Estimated time:** approximately 1 hour.

## How to use this worksheet

1. Open R or RStudio with `hello-eco-stats-r` as your working directory.
2. Source the launcher script to start **swirl**:

```{r}
source("01_introduction_to_r_programming.R")
```

3. In swirl, choose **Intro_Ecological_Statistics_in_R** →
   **01_introduction_to_r_programming**.
4. Work through each question below in the swirl console. Use this document as
   your reference — copy each question into the R source pane, write your
   answer beneath it, and run it.

---

## Section 1: R Basics & Vectors

### Background

Everything in R is stored as an **object**. The simplest objects hold a single
value (a scalar), while **vectors** hold a sequence of values of the same type.
You create objects with the **assignment operator** `<-` and combine values into
vectors with the `c()` function.

Common object types:

| Type        | Example                          |
|-------------|----------------------------------|
| Numeric     | `n_sites <- 4`                   |
| Character   | `region <- "Temperate woodland"` |
| Logical     | `is_wet <- TRUE`                 |
| Vector      | `days <- c("Mon", "Wed", "Fri")` |

### Questions

**Q1.** Which object type is the basic building block for storing a sequence of
values in R?

- vector
- plot
- package
- function

**Q2.** Create a character object called `study_region` and assign it the value
`"Temperate woodland"`.

```{r}
study_region <- "Temperate woodland"
```

**Q3.** Create a numeric object called `n_sites` with value `4`.

```{r}
n_sites <- 4
```

**Q4.** Create a vector called `survey_days` containing `"Mon"`, `"Wed"`, and
`"Fri"`.

```{r}
survey_days <- c("Mon", "Wed", "Fri")
```

**Q5.** Print `survey_days` to the console.

```{r}
survey_days
```

---

## Section 2: Reading & Inspecting Data

### Background

In ecological research, data usually lives in files — most commonly **CSV**
(comma-separated values) files. The base R function `read.csv()` reads a CSV
file into a **data frame**, R's standard tabular data structure. Once loaded,
you can preview data with `head()` (first 6 rows) and examine its structure with
`str()` (column names, types, dimensions).

A file called `eco_survey.csv` has been placed in your working directory. It
contains 12 rows of species count data from field surveys across wetland, forest
and grassland habitats.

| Column        | Description                            |
|---------------|----------------------------------------|
| `site`        | Site code (e.g. Wetland_A, Forest_C)   |
| `habitat`     | Habitat type                           |
| `species`     | Species observed                       |
| `count`       | Number of individuals recorded         |
| `temp_c`      | Air temperature (°C)                   |
| `rainfall_mm` | Rainfall (mm)                          |
| `observer`    | Field observer name                    |

### Questions

**Q6.** Read the file `"eco_survey.csv"` into an object called `eco_survey`
using `read.csv()`.

```{r}
eco_survey <- read.csv("eco_survey.csv")
```

**Q7.** Display the first 6 rows of `eco_survey` using `head()`.

```{r}
head(eco_survey)
```

**Q8.** Check the structure of `eco_survey` using `str()`.

```{r}
str(eco_survey)
```

**Q9.** Looking at the output of `str()`, how many observations (rows) are in
`eco_survey`?

- 12
- 7
- 4
- 24

**Q10.** In this dataset, what does the `count` column represent?

- Number of individuals observed
- Site elevation in meters
- Hours spent in the lab
- A species ID code

---

## Section 3: Basic Calculations & Subsetting

### Background

Once data is loaded you often need quick summaries. Key base R functions include:

- `mean()`, `sum()`, `median()` — descriptive statistics.
- `round(x, digits)` — round a number.
- `$` — access a single column (e.g. `eco_survey$count`).
- Logical indexing — subset rows with a condition
  (e.g. `df[df$col == "value", ]`).
- `nrow()` — count rows.

### Questions

**Q11.** Calculate the mean of the `count` column and save it as `mean_count`.

```{r}
mean_count <- mean(eco_survey$count)
```

**Q12.** Round `mean_count` to 2 decimal places.

```{r}
round(mean_count, 2)
```

**Q13.** Create `frog_rows` by keeping only rows where species is `"frog"` using
base R indexing.

```{r}
frog_rows <- eco_survey[eco_survey$species == "frog", ]
```

**Q14.** How many rows are in `frog_rows`? Use `nrow()`.

```{r}
nrow(frog_rows)
```

---

## Section 4: Introduction to the Tidyverse

### Background

The **tidyverse** is a suite of R packages (`dplyr`, `ggplot2`, `tidyr`,
`readr`, etc.) designed for intuitive data manipulation and visualisation. A
central feature is the **pipe operator** `%>%` (or the base R pipe `|>`), which
passes the output of one function into the first argument of the next — making
multi-step code read from left to right like a sentence.

```
data %>% step1() %>% step2() %>% step3()
```

Core dplyr verbs you will use today:

| Verb          | Purpose                          |
|---------------|----------------------------------|
| `select()`    | Choose columns                   |
| `filter()`    | Keep rows matching a condition   |
| `mutate()`    | Add or modify columns            |
| `group_by()`  | Group rows for summarisation     |
| `summarise()` | Compute summary statistics       |
| `arrange()`   | Sort rows                        |
| `count()`     | Tally observations               |

### Questions

Load the tidyverse package (swirl will prompt you).

```{r}
library(tidyverse)
```

**Q15.** Why do we use the pipe operator (`%>%`) in tidyverse workflows?

- It connects steps in a readable sequence
- It makes code run only on Windows
- It replaces all statistical tests
- It is required to create plots

Use `select()` and `head()` in a pipeline to view the `site`, `species`, and
`count` columns of `eco_survey` (swirl demo — not numbered).

```{r}
eco_survey %>% select(site, species, count) %>% head()
```

---

## Section 5: Filtering & Selecting Data

### Background

`filter()` keeps rows that match a logical condition (similar to a WHERE clause
in SQL). You can combine conditions with `&` (and) or `|` (or). `count()` is a
shortcut for grouping and tallying.

### Questions

**Q16.** Create `wetland_data` by filtering `eco_survey` to
`habitat == "wetland"`.

```{r}
wetland_data <- eco_survey %>% filter(habitat == "wetland")
```

**Q17.** Print `wetland_data`.

```{r}
wetland_data
```

**Q18.** Count how many records you have for each species, sorted from highest
to lowest.

```{r}
eco_survey %>% count(species, sort = TRUE)
```

---

## Section 6: Creating New Variables

### Background

`mutate()` creates new columns (or overwrites existing ones) inside a pipeline.
This is common when you need to **standardise** a variable — for example,
expressing counts per unit of rainfall to account for varying conditions across
surveys.

### Questions

**Q19.** Add a new column called `count_per_mm_rain = count / rainfall_mm` and
overwrite `eco_survey`.

```{r}
eco_survey <- eco_survey %>% mutate(count_per_mm_rain = count / rainfall_mm)
```

**Q20.** Show `species` and `count_per_mm_rain` for the first 6 rows.

```{r}
eco_survey %>% select(species, count_per_mm_rain) %>% head()
```

---

## Section 7: Summarising Data

### Background

`group_by()` + `summarise()` is the tidyverse equivalent of a pivot table.
You split the data into groups (e.g. by site or habitat) and compute summary
statistics within each group. Common summary functions include `mean()`,
`sum()`, `sd()`, `n()`, `min()`, and `max()`.

### Questions

**Q21.** Create `site_totals` with total individuals per site using `group_by()`
and `summarise()`.

```{r}
site_totals <- eco_survey %>%
  group_by(site) %>%
  summarise(total_count = sum(count))
```

**Q22.** Print `site_totals`.

```{r}
site_totals
```

**Q23.** Compute mean temperature by habitat (call the column `mean_temp`).

```{r}
eco_survey %>%
  group_by(habitat) %>%
  summarise(mean_temp = mean(temp_c))
```

**Q24.** Display the top 3 highest counts with `site`, `species`, and `count`.

```{r}
eco_survey %>%
  arrange(desc(count)) %>%
  select(site, species, count) %>%
  head(3)
```

---

## Section 8: Visualisation with ggplot2

### Background

**ggplot2** builds plots using a layered grammar of graphics. You start with
`ggplot(data, aes(...))` to set up the data and axis mappings, then add
**geoms** (geometric objects) such as `geom_point()`, `geom_boxplot()`,
`geom_line()`, etc. In ecology, boxplots are routinely used to compare abundance
distributions across habitat types or treatments.

### Questions

**Q25.** Create a boxplot of `count` by `habitat` using ggplot2.

```{r}
ggplot(eco_survey, aes(x = habitat, y = count)) +
  geom_boxplot()
```

> *Ecological interpretation:* notice how the median and spread of counts differ
> between habitats — this kind of visual comparison is often the first step
> before formal statistical testing.

---

## Section 9: Putting It All Together

### Background

Real ecological analyses typically chain several verbs together in a single
pipeline: group, summarise, arrange, then export. Practise combining the skills
you have learned so far.

### Questions

**Q26.** Create `species_summary` with `total_count` and `n_records` for each
species, sorted descending by `total_count`.

```{r}
species_summary <- eco_survey %>%
  group_by(species) %>%
  summarise(total_count = sum(count), n_records = n()) %>%
  arrange(desc(total_count))
```

**Q27.** Print `species_summary`.

```{r}
species_summary
```

**Q28.** Write `species_summary` to a CSV file named `"species_summary.csv"`
(without row names).

```{r}
write.csv(species_summary, "species_summary.csv", row.names = FALSE)
```

---

## Section 10: Reproducibility

### Background

Reproducibility is a cornerstone of good science. Recording which software
versions you used ensures that others (and future-you) can re-run your analysis
and obtain the same results. `sessionInfo()` prints the R version and all
loaded packages with their version numbers.

### Questions

**Q29.** Which action best supports reproducible ecological analysis?

- Keeping all analysis steps in scripts and re-running them
- Editing numbers manually in spreadsheets only
- Using different names for the same variable each day
- Running code once and not saving it

**Q30.** Run `sessionInfo()` so you can record software and package versions.

```{r}
sessionInfo()
```

---

## Summary

In this workshop you learned to:

1. Create and inspect R objects and vectors.
2. Read ecological survey data from a CSV file.
3. Calculate summary statistics and subset data with base R.
4. Build readable analysis pipelines with `%>%` and tidyverse verbs.
5. Produce exploratory boxplots with ggplot2.
6. Export tidy results to CSV.
7. Record your session for reproducibility.

**Next week:** data wrangling and exploratory ecological analysis on larger field
monitoring datasets.
